% =============================================================================
%                              MODULE OVERVIEW
% =============================================================================
\section*{Module Overview}
\addcontentsline{toc}{section}{Module Overview}

\subsection*{About This Module}

This module provides students with a comprehensive understanding of the fundamental concepts and techniques involved in \textbf{interpreting and extracting information from visual data}.

\textbf{Module code:} 6CCSAACV \hfill \textbf{Semester:} 2 (AY 2025/26)

\textbf{Module Lecturer:} Dr.\ Luis Carlos Garcia Peraza Herrera

\bigseparator

\subsection*{Module Aims}

The module aims to:
\begin{itemize}
  \item Provide a comprehensive understanding of fundamental concepts and techniques in computer vision.
  \item Develop skills in interpreting and extracting information from visual data.
  \item Cover both traditional and state-of-the-art machine learning techniques for vision tasks.
\end{itemize}

\bigseparator

\subsection*{Learning Outcomes}

By the end of this module, you will be able to:

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{@{} c p{12cm} @{}}
\toprule
\textbf{Code} & \textbf{Learning Outcome} \\
\midrule
ACV1 & Identify and solve complex computer vision tasks, selecting and implementing the most suitable algorithms and methodologies. \\
ACV2 & Design and develop innovative methods and algorithms for addressing various computer vision challenges, critically evaluating their strengths and limitations. \\
ACV3 & Analyse, assess, and effectively deploy a range of traditional and state-of-the-art machine learning techniques to implement robust solutions to diverse computer vision tasks. \\
\bottomrule
\end{tabular}
\end{center}

\bigseparator

\subsection*{Recommended Reading}

There is no required reading for this module. However, the following books are recommended:

\begin{enumerate}
  \item \textit{Computer Vision: Algorithms and Applications} (2nd ed). Richard Szeliski. Springer, 2022.
  \item \textit{Seeing: The Computational Approach to Biological Vision} (2nd ed). John P.\ Frisby and James V.\ Stone. MIT Press, 2010.
  \item \textit{Computer Vision: Models, Learning, and Inference}. Simon J.\ D.\ Prince. Cambridge University Press, 2012.
  \item \textit{Deep Learning: Foundations and Concepts}. Christopher Bishop. Springer, 2023.
\end{enumerate}

\bigseparator

\subsection*{Assessment}

\TODO{} Add assessment details when provided (coursework + exam breakdown).

\bigseparator

\subsection*{Weekly Topics}

\TODO{} Add weekly topic overview once lecture slides are provided.
